Relationships in the extremes and their influence on competition and coexistence

### 2022-01-09 paper/Paper.Rnw
edits to Intro

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Descamps_2005, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}A,C 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}A, termed ``left-tail association'') or right (Fig. \ref{fig:noise}C,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

%Intro to storage effects and why they might be influenced by ATAs
Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. 
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can or may alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species. Although we focus here on temporal coexistence
mechanisms, ATAs between environmental variables measured in different places may also be altered 
by climate change, if more intensive climatic events are also more spatially extensive.

@@

@@

Fig.\ref{fig:fig6} displays the subsets of environmental parameters for which coexistence occurs or fails and the degree of ATA contribution to the IGR. Specific to this model and parameter ranges, because $\Delta_i^{[EC]}$ is always positive, all positive values of $\Delta_i^{[EC]}/\text{IGR}$ indicate coexistence and negative values of $\Delta_i^{[EC]}/\text{IGR}$ indicate failure of coexistence. Futhermore, ATA exlusion does not occur and ATA rescue occurs in positive cases where $\Delta_i^{[EC]}>IGR$. To preserve simplicity and readibility of the plots, $|\Delta_i^{[EC]}/\text{IGR}>1|$ were capped at $|1|$. Higher magnitudes of $\Delta_i^{[EC]}/\text{IGR}$ tended to occur as the IGR became smaller and smaller, thus creating the sharp contrast in the plots where highest intensity positive spaces quickly change into highest intensity negative spaces with decreasing parameters. Like the result of Fig.\ref{fig:fig5}, we see in Fig.\ref{fig:fig6} that amplitude and mean temperature have the most influence on coexistence. Coexistence occurs as $\theta_0$ and $a$ increase and $\Delta_i^{[EC]}/\text{IGR}$ changes from negative to positive. The effect of varying periodicity of environmental fluctuations is...

@@

\begin{figure}
\includegraphics[width=.25\textwidth]{../results_figs/fig1_vertical.pdf}
\caption{Simple example of bivariate random variables with standard normal marginals showing
alternative patterns of association in the tails. A) is ``left-tail 
associated,'' B) is symmetric, and C) is ``right-tail associated'' noise. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, yet show very different 
patterns of association. We used asymmetrically tail associated cases which are perfectly 
correlated in the lower (A)  or upper (C) halves of the distributions, though it is also 
possible to generate asymmetrically tail associated noise with imperfect correlations in 
both halves of the distributions \citep{Ghosh_2020_AER,Nelsen2006}. The variables $(b_1,b_2)$ described in
the Introduction and Methods were obtained by generating 
noise as pictured and then multipling by $\sigma$ and adding $(\mu_1,\mu_2)$.\label{fig:noise}}
\end{figure}
%***DAN: Dan has edited this caption at least once.

\begin{figure}
\includegraphics[width=.75\textwidth]{../results_figs/fig2.pdf}
\caption{Our simulated lottery-model populations influenced by asymmetric tail associations (ATAs) recovered from low densities less often and 
spent more time in a state of one-species dominance, demonstrating that ATAs can influence aspects of 
species coexistence. Panel A shows species 1, from the lottery 
model subject to environmental noise with symmetric tail associations (Fig. \ref{fig:noise}B). Panel B shows species 1 from the 
same model with the same parameters, but subject to noise with left-tail association (Fig. \ref{fig:noise}A).
Both simulations were run for \Sexpr{M} time steps, though only the first $1000$ are shown. Periods of 
``noticeable coexistence'' of the species, defined to be periods when both species each comprise at least $5\%$ of 
total abundance, are marked in grey. The average duration of periods of noticeable coexistence 
was \Sexpr{fig2$coSym} time steps for panel A and \Sexpr{fig2$coATA} for panel B, over the full length of the simulations; 
whereas the average duration of periods of
one-species dominance was \Sexpr{fig2$domSym} time steps for panel A and \Sexpr{fig2$domATA} for B. Thus, while periods of 
noticeable coexistence were of similar duration for the two scenarios, periods of one-species dominance were much longer,
on average, for the ATA case. Parameters were $\mu_1 = \Sexpr{fig2$params[1] + fig2$params[1]}$ and $\mu_2 = \Sexpr{-(fig2$params[1])}$, $\sigma = \Sexpr{fig2$params[2]}$, and $\delta = \Sexpr{fig2$params[3]}$. See Introduction for model details. 
\label{fig:pedagog}}
\end{figure}
%***DAN: Dan has edited this caption

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig3.pdf}
\caption{The contribution of ATAs to species coexistence is often meaningfully large compared to 
other components of the MCT decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and the IGR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs (red line) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the IGR line (orange) falls 
between the ATA line (red)
and the $x$ axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the weaker competitor. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) was generated 
using $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. Though results differed
in some details, main conclusions were the same. \label{fig:fig3}}
\end{figure}
%***DAN: Dan has edited this caption

>>>>

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Descamps_2005, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
%***DAN: Jasmin, Descamps_2005 does not appear in refs.bib, I think. Please add it there or subtract the callout here, whatever is most appropriate.
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}a,c 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}a, termed ``left-tail association'') or right (Fig. \ref{fig:noise}c,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

%Intro to storage effects and why they might be influenced by ATAs
Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. 
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species are currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017, Lyon2019, Keelings2020}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species. Although we focus here on temporal coexistence
mechanisms, ATAs between environmental variables measured in different places may also be altered 
by climate change, if more intensive climatic events are also more spatially extensive.

@@

%***DAN: Dan made it to here on 2022 01 09 editing, continue tomorrow.

@@

Fig.\ref{fig:fig6} displays the subsets of environmental parameters for which coexistence occurs or fails and the degree of ATA contribution to the IGR. Specific to this model and parameter ranges, because $\Delta_i^{[EC]}$ is always positive, all positive values of $\Delta_i^{[EC]}/\text{IGR}$ indicate coexistence and negative values of $\Delta_i^{[EC]}/\text{IGR}$ indicate failure of coexistence. Futhermore, ATA exlusion does not occur and ATA rescue occurs in positive cases where $\Delta_i^{[EC]}>IGR$. To preserve simplicity and readibility of the plots, $|\Delta_i^{[EC]}/\text{IGR}>1|$ were capped at $|1|$. Higher magnitudes of $\Delta_i^{[EC]}/\text{IGR}$ tended to occur as the IGR became smaller and smaller, thus creating the sharp contrast in the plots where highest intensity positive spaces quickly change into highest intensity negative spaces with decreasing parameters. Like the result of Fig. \ref{fig:fig5}, we see in Fig. \ref{fig:fig6} that amplitude and mean temperature have the most influence on coexistence. Coexistence occurs as $\theta_0$ and $a$ increase and $\Delta_i^{[EC]}/\text{IGR}$ changes from negative to positive. The effect of varying periodicity of environmental fluctuations is...

@@

\begin{figure}
\includegraphics[width=.25\textwidth]{../results_figs/fig1_vertical.pdf}
\caption{Simple example of bivariate random variables with standard normal marginals showing
alternative patterns of association in the tails. The varables in (a) are ``left-tail 
associated,'' those in (b) are symmetrically associated, and those in (c) are ``right-tail associated''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these very different 
patterns of association. We used asymmetrically tail associated cases (a, c) which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also 
possible to generate asymmetrically tail associated noise with imperfect correlations in 
both halves of the distributions \citep{Ghosh_2020_AER,Nelsen2006}. The variables $(b_1,b_2)$ described in
the Introduction and Methods were obtained by generating 
noise as pictured and then multipling by $\sigma$ and adding $(\mu_1,\mu_2)$.\label{fig:noise}}
\end{figure}
%***DAN: Jasmin, I appologize for the following pedantic request, I explain below why it is necessary. Can you please change panel
%labels on this and all figure from the format "A", "B", "C" to the format "(a)", "(b)", "(c)" (i.e., lower case panel labels
%with parentheses around them). The reason for this is that this is Ecology Letters style. If we do not follow their style then their
%copy editors will make the change for us. I would prefer not to have copy editors messing around with our figures. Though Ecology Letters
%copy editors are usually pretty good, I have had many bad experiences with copy editors, generally, and I have had at least a few bad
%experiences with Ecology Letters copy editors. What happens is, they introduce errors into a paper while typesetting it, and one has to
%check their work very carefully to avoid those mistakes going to press. It's hard to blame them since it must be the most boring job 
%in the world, but anyway we do not want to have errors in our published paper as a result. The best way to prevent them introducing 
%errors into our figures is to not given them any reason to make any changes at all to our figures. The same change has to be 
%made to all main text and sup mat figures, and then all callouts to figures have to also be so changed, please. I will do these
%as much as possible while I edit, but if you could please also check it will help, and I am relying on you to change the panel labels
%in the figured themselves. I am sorry again for the tedium, here. Maybe some day they will improve the publication process so we do not 
%have to worry about nonsense like this.

\begin{figure}
\includegraphics[width=.75\textwidth]{../results_figs/fig2.pdf}
\caption{Our simulated lottery-model populations influenced by asymmetric tail associations (ATAs) recovered from low densities less often and 
spent more time in a state of one-species dominance, demonstrating that ATAs can influence aspects of 
species coexistence. Panel (a) shows species 1, from the lottery 
model subject to environmental noise with symmetric tail associations (Fig. \ref{fig:noise}b). Panel (b) shows species 1 from the 
same model with the same parameters, but subject to noise with left-tail association (Fig. \ref{fig:noise}a).
Both simulations were run for \Sexpr{M} time steps, though only the first $1000$ are shown. Periods of 
``noticeable coexistence'' of the species, defined to be periods when both species each comprise at least $5\%$ of 
total abundance, are marked in grey. The average duration of periods of noticeable coexistence 
was \Sexpr{fig2$coSym} time steps for panel (a) and \Sexpr{fig2$coATA} for panel (b), over the full length of the simulations; 
whereas the average duration of periods of
one-species dominance was \Sexpr{fig2$domSym} time steps for panel (a) and \Sexpr{fig2$domATA} for (b). Thus, while periods of 
noticeable coexistence were of similar duration for the two scenarios, periods of one-species dominance were much longer,
on average, for the ATA case. Parameters were $\mu_1 = \Sexpr{fig2$params[1] + fig2$params[1]}$ and $\mu_2 = \Sexpr{-(fig2$params[1])}$, $\sigma = \Sexpr{fig2$params[2]}$, and $\delta = \Sexpr{fig2$params[3]}$. See Introduction and Fig. \ref{fig:noise} caption for model details. 
\label{fig:pedagog}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig3.pdf}
\caption{The contribution of ATAs to species coexistence is often meaningfully large compared to 
other components of the MCT decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MCT, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and the IGR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs (red line) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the IGR line (orange) falls 
between the ATA line (red)
and the $x$ axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the weaker competitor. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) was generated 
using $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. Though results differed
in some details, main conclusions were the same. \label{fig:fig3}}
\end{figure}
